Natural variation in teosinte at the domestication locus \emph{teosinte branched1} (\emph{tb1})

### 2014-05-20 manuscript.tex
JRI edits to intro, typos in refs

The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of lateral organ growth, is a major QTL involved in branching differences between maize and its wild progenitor, teosinte. Further studies have shown that the insertion of a transposable element (\emph{Hopscotch}) upstream of \emph{tb1} enhances its expression, causing the reduction in branching observed in domesticated maize. Observations of the maize \emph{tb1} allele in teosinte individuals, coupled with estimates of the age of insertion of the \emph{Hopscotch} element, led us to investigate the prevalence and potential role of \emph{tb1} in teosinte. Results from genotyping across many natural populations suggest that the \emph{Hopscotch} element is segregating at a higher than expected frequency in a number of populations of two subspecies of teosinte, \emph{Zea mays} ssp. \emph{parviglumis} and \emph{Zea mays} ssp. \emph{mexicana}. Analysis of linkage disequilibrium between the \emph{Hopscotch} element and variation in surrounding regions does not support a hypothesis of recent introgression from maize into teosinte, and we find no evidence of environmental correlations that might suggest recent selection. Finally, two greenhouse experiments with \emph{Zea mays} ssp. \emph{parviglumis} do not suggest an important role for \emph{tb1} in controlling tillering in natural populations of this subspecies. Our findings suggest that the role of the \emph{Hopscotch} in tillering in teosinte is not as straightforward as is in domesticated maize, and that other loci may play a role in observed variation in this trait.

@@

Domesticated crops and their wild progenitors provide an excellent system in which to study adaptation and genomic changes associated with human-mediated selection \citep{Ross-Ibarra2007}. Perhaps the central focus of the study of domestication has been the identification of genetic variation underlying agronomically important traits such as fruit size and plant architecture \citep{OlsenGross2010}. Additionally, many domesticates show reduced genetic diversity when compared to their wild progenitors, and an understanding of the distribution of diversity in the wild and its phenotypic effects has become increasingly useful to crop improvement \citep{KovachMcCouch2008}. But while some effort has been invested into understanding how wild alleles behave in their domesticated relatives \citep{Bai2007}, very little is known about the role that alleles found most commonly in domesticates play in natural populations of their wild progenitors.
\citep{Whitton1997}.

@@

Many morphological changes are associated with maize domestication, and understanding the genetic basis of these changes has been a focus of maize research for a number of years \citep{Doebley2004}. One of the most dramatic changes is found in plant architecture: domesticated maize is characterized by a central stalk with few tillers and lateral branches terminating in a female inflorescence, while teosinte is highly tillered and bears tassels (male inflorescences) at the end of its lateral branches. The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of organ growth, was identified as a major QTL involved in domestication branching differences achieved through higher expression of the maize allele than the teosinte allele \citep{DoebleyStecGustus1995, Doebley1997}. Further work showed that the insertion of a 4.9 kb retrotransposon (\emph{Hopscotch}) in the upstream control region of \emph{tb1} caused the increased expression of this gene reduction in branching observed in domesticated maize \cite{Studer2011}. The effects of this insertion have been observed in tiller number in maize, but little is known about its role, if any, in natural populations of teosinte \citep{Studer2011}. Dating of this element has suggested that its insertion predates the domestication of maize, leading to the hypothesis that it was segregating as standing variation in ancient populations of teosinte and increased to high frequency in maize due to selection during domestication \citep{Studer2011}. Furthermore, \citet{StuderDoebley2012} investigated the phenotypic effects of nine teosinte \emph{tb1} alleles in an isogenic maize background and found that the introgressions sort into three distinct phenotypic classes, suggesting that variation at the \emph{tb1} locus may play a functional role in teosinte.

In high-density species such as teosinte, plants can detect impending competition from their neighbors through detection of the ratio of red to far-red light. An increase in far-red relative to red light accompanies shading and triggers physiological and morphological changes such as reduced tillering, increased plant height and early flowering collectively known as the shade avoidance syndrome \citep{KebromBrutnell2007}. The \emph{tb1} locus appears to play an important role in the shade avoidance pathway in \emph{Zea mays} and other grasses and may therefore be crucial to the ecology of teosinte \citep{KebromBrutnell2007, LukensDoebley1999}.  In this study we aim to characterize the distribution of the \emph{Hopscotch} insertion in \emph{parviglumis}, \emph{mexicana}, and landrace maize, and to examine the phenotypic effects of the insertion in \emph{parviglumis}. We use a combination of PCR genotyping for the \emph{Hopscotch} element in our full panel and sequencing of two small regions upstream of \emph{tb1} in a subset of teosinte populations to explore patterns of genetic variation at this locus. Finally, we test for an association between the \emph{Hopscotch} element and tillering phenotypes in a population of \emph{parviglumis}.

>>>>

The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of lateral organ growth, is a major QTL involved in branching differences between maize and its wild progenitor, teosinte. Previous work has shown that the insertion of a transposable element (\emph{Hopscotch}) upstream of \emph{tb1} enhances its expression, causing much of the reduction in tillering observed in domesticated maize. Observations of the maize \emph{tb1} allele in teosinte individuals, coupled with estimates of the age of insertion of the \emph{Hopscotch} element, led us to investigate the prevalence and potential role of \emph{tb1} in teosinte. Results from genotyping across many natural populations suggest that the \emph{Hopscotch} element is segregating at a higher than expected frequency in a number of populations of two subspecies of teosinte, \emph{Zea mays} ssp. \emph{parviglumis} and \emph{Zea mays} ssp. \emph{mexicana}. Analysis of linkage disequilibrium between the \emph{Hopscotch} element and variation in surrounding regions does not support a hypothesis of recent introgression from maize into teosinte, and we find no evidence of environmental correlations that might suggest recent selection. Finally, two greenhouse experiments with \emph{Zea mays} ssp. \emph{parviglumis} do not suggest an important role for \emph{tb1} in controlling tillering in natural populations of this subspecies. Our findings suggest that the role of the \emph{Hopscotch} in tillering in teosinte is not as straightforward as is in domesticated maize, and that other loci may play a role in observed variation in this trait.

@@

Domesticated crops and their wild progenitors provide an excellent system in which to study adaptation and genomic changes associated with human-mediated selection \citep{Ross-Ibarra2007}. Perhaps the central focus of the study of domestication has been the identification of genetic variation underlying agronomically important traits such as fruit size and plant architecture \citep{OlsenGross2010}. Additionally, many domesticates show reduced genetic diversity when compared to their wild progenitors, and an understanding of the distribution of diversity in the wild and its phenotypic effects has become increasingly useful to crop improvement \citep{KovachMcCouch2008}. But while some effort has been invested into understanding how wild alleles behave in their domesticated relatives \citep[e.g.][]{Bai2007}, very little is known about the role that alleles found most commonly in domesticates play in natural populations of their wild progenitors \citep{Whitton1997}.

@@

Many morphological changes are associated with maize domestication, and understanding the genetic basis of these changes has been a focus of maize research for a number of years \citep{Doebley2004}. One of the most dramatic changes is found in plant architecture: domesticated maize is characterized by a central stalk with few tillers and lateral branches terminating in a female inflorescence, while teosinte is highly tillered and bears tassels (male inflorescences) at the end of its lateral branches. The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of organ growth, was identified as a major QTL involved in  branching \citep{DoebleyStecGustus1995} and tillering \citep{DoebleyStec1991} differences between maize and teosinte.  A 4.9 kb retrotransposon (\emph{Hopscotch}) insertion into the upstream control region of \emph{tb1} in maize acts to enhance expression of \emph{tb1}, thus repressing lateral organ growth \citep{Doebley1997, Studer2011}.  Dating of the \emph{Hopscotch} retrotransposon suggests that its insertion predates the domestication of maize, leading to the hypothesis that it was segregating as standing variation in ancient populations of teosinte and increased to high frequency in maize due to selection during domestication \citep{Studer2011}. The effects of the \emph{Hopscotch} insertion have been studied in maize \citep{Studer2011}, and analysis of teosinte alleles at \emph{tb1} has identified functionally distinct allelic classes \citet{StuderDoebley2012}, but little is known about the role of \emph{tb1} or the \emph{Hopscotch} insertion in natural populations of teosinte.

In teosinte and other plants that grow at high population density, individuals detect competition from neighbors via the ratio of red to far-red light.  An increase in far-red relative to red light accompanies shading and triggers the shade avoidance syndrome: a suite of physiological and morphological changes such as reduced tillering, increased plant height and early flowering \citep{KebromBrutnell2007}. The \emph{tb1} locus appears to play an important role in the shade avoidance pathway in \emph{Zea mays} and other grasses and may therefore be crucial to the ecology of teosinte \citep{KebromBrutnell2007, LukensDoebley1999}.  In this study we aim to characterize the distribution of the \emph{Hopscotch} insertion in \emph{parviglumis}, \emph{mexicana}, and landrace maize, and to examine the phenotypic effects of the insertion in \emph{parviglumis}. We use a combination of PCR genotyping for the \emph{Hopscotch} element in our full panel and sequencing of two small regions upstream of \emph{tb1} in a subset of teosinte populations to explore patterns of genetic variation at this locus. Finally, we test for an association between the \emph{Hopscotch} element and tillering phenotypes in a population of \emph{parviglumis}.
